General linear Models Coursework

### 2022-11-22 Draft_final.Rmd
Going over and slightly rewording inference

knitr::kable(Model.comparison.data,digits=3,caption = "Comparing the AIC and adjusted R squared of models from part 2-5")%>%
  row_spec(0,bold=T)%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```
The model from part 5, $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\beta_{2}x^2_{i}+\epsilon$ has the lowest AIC value of **1528.062** and the highest adjusted R squared value of **0.4635055**. Furthermore, from the diagnostic plots the normality assumption is valid and the fitted vs residual plot displays an even scattering of points around 0. Therefore the 4th model is my preferred model to describe the relationship between expenditure and income.

The log-linear model means that the covarites (income and income squared) have a multiplicative effect on the response (expenditure). The squared term of income included in the model, means that the change in the mean of expenditure depends on the value of income. Income has a positive association with expenditure with an increase in income causing a percent increase in expenditure. With the initial value of income affecting the magnitude of the percent increase on expenditure. Smaller initial values of income cause a greater percent increase on expenditure (due to the squared income term being negative). In particular, using a realistic increase in weekly income of £20 for a household earning the mean income of £512 gives an increase of $(\frac{exp(532\beta_1+532^{2}\beta_{2})}{exp(512\beta_1+512^{2}\beta_{2})}-1)$x100% = **3.25%** in expenditure.

```{r,include=F}
#calculating percent increase
beta_1 <- 3.117180e-03	
beta_2 <- -1.453272e-06	
x <- 512
inc <- 20
#Formual to calculate oercent increase
(exp(beta_1*(x+inc)+beta_2*(x+inc)^2)/exp(beta_1*x+beta_2*x^2)-1)*100
```

@@

Due to the model being a log-linear model the explanatory variables have a multiplicative effect on expenditure. Exponentiating the estimated coefficients to find their effect on expenditure gives the below table. I note that all values have been rounded to 4 decimal places thus the zero p-values are due to p-values being too small to be displayed and similarly for effect of income squared. My following inference will be based on this table and I use a significance level of 5%.

@@

```
The difference between full time work and being unemployed was highly statistically significant. With unemployed households compared to full time employment households having a 95% confidence interval (CI) of [0.659, 0.915] which corresponds to a **8.5%** to **34.1%** decrease in expenditure with all other variables held constant. Furthermore, as the CI of unemployed and part-time employment do not overlap there is a statistically significant increase in expenditure if a household is in part time employment compared to unemployed households with all other variables held constant.

However the difference between full time work and being inactive or in part time work was not significant (the 95% CI included 1) therefore with all other variables constant there is no evidence that being in one of these three employment groups compared to another influences the expenditure of the household.

The difference between a households expenditure for households which live in a publicly rented property and either a privately rented or owned property were statistically significant. With the associated 95% CI of [1.22, 1.441] and [1.115 , 1.281] respectively which corresponds to a **22%** to **44.1%** and a **11.5%** to **28.1%** increase in expenditure respectively with all other variables held constant.

Sex of household head and household size have an interaction between them therefore cannot be reported separately. To easily display the effects of both on expenditure I created the below table which displays the multiplicative effect of sex and household size on expenditure.

@@

For households of size 1,2 and 4 a female household head corresponds to an increase in expenditure compared to a male household of size 1, however for a household of size 3 or 5+ there is no evidence of a difference in expenditure (the CI includes 1). I note the confidence intervals are very large for female headed households of size 2 and above, this means there is alot of variation in the data causing the uncertainty in the results. This could be due to the small sample of female household heads of size 3 and above (insufficient power) or additional confounders affecting female household heads that were not included in the analysis.

Compared to a household of size 1, all larger sized households increase expenditure if the household head is male with an increase of expenditure of **22.4%** to **45.5%**, **36.2%** to **71.9%**, **31.3%** to **66.8%** and **34.9%** to **81.6%** for households of size 2, 3, 4 and 5+ respectively with all other variables held constant. There is no evidence larger households increase expenditure for female headed households as the 95% CI's overlap for all household sizes.

Due to income having a squared term the effect of an increase in income on expenditure depends on the initial value of income. To demonstrate this I produced the plot below showing how the percent increase in expenditure varies with the initial income of a household for several different increases in household income. For all but the most extreme household incomes (slightly over £1000), income has a positive effect on expenditure. With the higher the initial income the less of a positive effect on expenditure an increase in income has (seen by negative gradient in below plot and which is due to the negative income squared regression coefficient). 
```{r,include=F}
#calculating percent increase
beta_1 <- 0.002285
beta_2 <- -0.000001036
x <- 900
inc <- 20

(exp(beta_1*(x+inc)+beta_2*(x+inc)^2)/exp(beta_1*x+beta_2*x^2)-1)*100

@@

Overall an increase in income (except for intially extremely high incomes), living in an owned or privately rented property compared to a publicly rented property and being in full or part time employment compared to unemployed increases the expenditure of the household. With household size increasing income if household head is male and being a female household head increases expenditure for certain household sizes (1, 2 and 4) and having no proven effect for others (3 and 5+).

It is important to remember that the relationships found in this analysis does not imply causation as they may be affected by confounding, chance or bias from the sample the model was based on. Including additional factors that could potentially be confounders such as the age of the household head, ethnicity of household and location of household would improve the confidence in the relationships found. Furthermore, a larger sample would enable the effect of sex of household head to be better understood for all household sizes.

>>>>

knitr::kable(Model.comparison.data,digits=3,caption = "Comparing the AIC and adjusted R squared of models from part 2-5")%>%
  row_spec(0,bold=T)%>%
  kable_styling(position = "center",latex_options = "HOLD_position")
```
The model from part 5, $log(y_{i})=\beta_{0}+\beta_{1}x_{i}+\beta_{2}x^2_{i}+\epsilon_{i}$ has the lowest AIC value of **1528.062** and the highest adjusted $R^{2}$ value of **0.464** thus both model selection criteria select model 5 as the best fitting model of the data. Furthermore, from the diagnostic plots the normality assumption is valid and the fitted vs residual plot displays an even scattering of points around 0. Therefore the 4th model from part 5 is my preferred model to describe the relationship between expenditure and income.

This is a log-linear model which means that the covarites (income and income squared) have a multiplicative effect on the response (expenditure). The squared term of income, means that the effect of income on expenditure depends on the initial value of income. Due to the negative value of the squared income regression coefficient the lower the initial value of income the greater the percent increase on expenditure for a given increase in income.  Overall, income has a positive association with expenditure in the model with an increase in income causing a percent increase in expenditure (for all but the most extreme initial values of income of slightly over £1060). In particular, using a realistic increase in weekly income of £20 for a household earning the mean income of £512 gives an increase of $(\frac{exp(532\beta_1+532^{2}\beta_{2})}{exp(512\beta_1+512^{2}\beta_{2})}-1)$x100% = **3.25%** in household expenditure.

```{r,include=F}
#calculating percent increase

model.p5.increase <- function(initial,increase){
  #values of regression coefficients
  beta_1 <- 3.117180e-03	
  beta_2 <- -1.453272e-06

 #Formula to calculate percent increase
 perc.inc <- (exp(beta_1*(initial+increase)+beta_2*(initial+increase)^2)/exp(beta_1*initial+beta_2*initial^2)-1)*100

 return(perc.inc)
}

model.p5.increase(512,20)

#finding when increase in expenditure becomes negative
model.p5.increase(1050,20)
model.p5.increase(1060,20)
model.p5.increase(1070,20)
```

@@

Due to the model being a log-linear model the explanatory variables have a multiplicative effect on expenditure. Exponentiating the estimated coefficients to find their effect on expenditure gives the below table. I note that all values have been rounded to 4 decimal places thus the zero p-values are due to p-values being too small to be displayed and similarly for effect of income squared. The exponentiated estimate of the income squared regression coefficient is 0.99999896 (8 d.p.) with a 95% confidence interval (CI) of [0.99999866, 0.99999927]. My following inference will be based on this table using a significance level of 0.05.

@@

```
The difference between households in full time employment and those unemployed was highly statistically significant. With unemployed households compared to full time employment households having an exponentiated 95% CI of [0.659, 0.915] which corresponds to a **8.5%** to **34.1%** decrease in expenditure with all other variables held constant. Furthermore, as the CI of unemployed and part-time employment do not overlap there is a statistically significant increase in expenditure if a household is in part time employment compared to unemployed households with all other variables held constant. However, there was no evidence to support a difference in expenditure between full time employment households and those that were inactive or in part time work (the exponentiated 95% CI included 1).

The difference between a households expenditure for households which live in a publicly rented property and either a privately rented or owned property were statistically significant. With an exponentiated 95% CI of [1.22, 1.441] and [1.115 , 1.281] respectively which corresponds to a **22%** to **44.1%** and a **11.5%** to **28.1%** increase in expenditure respectively with all other variables held constant.

Sex of household head and household size have an interaction between them therefore cannot be reported separately. To easily display the effects of both on expenditure simultaneously I created the below 2x2 table which displays the multiplicative effect of sex and household size on expenditure.

@@

For households of size 1,2 and 4 a female household head corresponds to an increase in expenditure compared to a male household of size 1, however for a household of size 3 or 5+ there is no evidence of a difference in expenditure (the exponentiated CI includes 1) with all other variables held constant. I note the confidence intervals are very large for female headed households of size 2 and above, this means there is alot of variation in the data causing this uncertainty in the results. This could be due to the small sample of female household heads of size 3 and above (insufficient power) or additional confounders affecting female household heads that were not included in the analysis.

Compared to a household of size 1, all larger sized households increase expenditure if the household head is male with an increase of expenditure of **22.4%** to **45.5%**, **36.2%** to **71.9%**, **31.3%** to **66.8%** and **34.9%** to **81.6%** for households of size 2, 3, 4 and 5+ respectively with all other variables held constant. There is no evidence larger households increase expenditure for female headed households as the 95% exponentiated CI's overlap for all household sizes.

Due to income having a squared term the effect of an increase in income on expenditure depends on the initial value of income. To demonstrate this I produced the plot below showing how the percent increase in expenditure varies with the initial income of a household for several different increases in household income. For all but the most extreme household incomes (slightly over £1000), income has a positive effect on expenditure. With the higher the initial incomes the less the percent increase in expenditure for the same increase in income (seen by negative gradient in below plot, due to the negative income squared regression coefficient). For an initial income of £512 (the mean income) an increase in income of £20 and £50 increases expenditure by **2.44%** and **6.04%** respectively.
```{r,include=F}
#calculating percent increase of expenditure from increasing income
model.final.increase <- function(initial,increase){
  #values of regression coefficients
  beta_1 <- 0.002285
  beta_2 <- -0.000001036

 #Formula to calculate percent increase
 perc.inc <- (exp(beta_1*(initial+increase)+beta_2*(initial+increase)^2)/exp(beta_1*initial+beta_2*initial^2)-1)*100

 return(perc.inc)
}

model.final.increase(512,20)
model.final.increase(512,50)

@@

Overall an increase in income (except for households on extremely high incomes), living in an owned or privately rented property compared to a publicly rented property and being in full or part time employment compared to unemployed increases the expenditure of a household. With household size increasing income if household head is male and being a female household head increases expenditure for certain household sizes (1, 2 and 4) and having no proven effect for others (3 and 5+).

It is important to remember that the relationships found in this analysis does not imply these variables cause an increase in expenditure as they may be affected by confounding, chance or bias from the sample the model was based on. However some areas of causal confidence are: (1) income, logically if you make more money you have more available to spend, (2) Employment status, if you are unemployed you are going to be more frugal with purchases. Including additional factors that could potentially be confounders such as the age of the household head, ethnicity of household and location of household would improve the confidence in the relationships found. Furthermore, a larger sample would enable the effect of sex of household head to be better understood for all household sizes.
